Stability in human interaction networks: primitive typology of vertex, prominence of measures and time activity statistics

### 2015-04-22 paper.tex
adicionadas versoes tex pdf e docx corretas

\section{Introduction}\label{sec:into}
Human interaction networks have been studied long before modern computers.
Antecedents can be found in nineteenth century works, while the foundation of
the social network analysis field is generally attributed to the psychiatrist Jacob Moreno~\cite{newmanBook}.
More recently, the continuous increase of stored data related to human interaction has stimulated research,
and overall characterization is scattered in literature.

Reports on complex networks are usually focused on a particular measure or functionality, such
as accessibility measure or detecting communities~\cite{access,newmanModularity}. Although there are research on networks arisen from specific phenomenon, such as social or protein networks, there are few attempts for a general characterization and even less endeavors that take in account the temporal dynamic.

The present work is aimed at finding common and stable characteristics among (email) interaction networks. This is achieved through observations along time, which imply network evolution, a field that has received attention from the research community
for more than a decade~\cite{barabasiEvo,newmanEvolving}. The evolution is represented herein by a timeline of activity snapshots with a constant number of contiguous messages. These snapshots yield a succession of networks. This approach, albeit coherent and intuitive, seems not to be explored to date~\cite{netWindowEvo}.

We consider only directed, weighted and human interaction networks. Undirected and unweighted representations of such networks are also found in the literature and can be obtained by simplification~\cite{GMANE2}.
Meaningful measures depend heavily on the model and system characteristics~\cite{newmanStru, newmanWeight}, therefore the most basic (and ubiquitous, and fundamental) metrics of centrality, clustering and symmetry are analyzed altogether in principal components formation.

Typologies are the canon of scientific literature for classification of human agents~\cite{typCanon}. Consequently, it was considered central to benefit from human typologies. This is a complex matter carefully addressed in methodology, results and appendix.

Although all networks considered originated from email lists,
 coherence with literature suggests that results hold for a more general class of interaction networks,
 such as observed in online platforms (e.g. LinkedIn, Facebook, Twitter).
Indeed, text mining routines and typologies of online participants benefit from the results here presented~\cite{rcText,rcTipo}.

Next subsection exposes related work. Section~\ref{sec:data} details data while Section~\ref{sec:carac} is dedicated to the characterization methods. Section~\ref{sec:results} holds results and discussions, leading to Section~\ref{sec:conc} for conclusions and further work. Appendix~\ref{ap:ded} develops simple graph results used in the methods, Appendix~\ref{scripts} exposes access to data and scripts, and Appendix~\ref{ap:typ} further introduces the typological approach. Detailed tables and figures are bundled in Support Information, and are cited upon need.

@@

The evolution of human interaction networks was addressed with a community focus, where the direction of edges was not taken into account~\cite{barabasiEvo}. In another article, two topologically different networks emerged from human interaction networks, depending on the frequency of interactions, which can either be a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. In email list networks, scale-free properties were reported with $\alpha=1$~\cite{bird} (as are web browsing and library loans~\cite{barabasiHumanDyn}), and different linguistic traces were related to weak and strong ties~\cite{GMANE2}.

Unreciprocated edges often exceed 50\% in the networks analyzed, which matches empirical evidence from literature~\cite{newmanEvolving}. No correlation of topological characteristics and geographical coordinates was found~\cite{barabasiGeo}, therefore geographical positions were discarded in the present article.
On the other hand, gender related behavior in mobile phone datasets has been reported~\cite{barabasiSex} and was not considered in this article as the email messages and addresses have no gender related metadata~\cite{GMANE}.

There are research reports based on the results presented in this article, which are exposed in Section~\ref{sec:conc}.

@@

Although many lists were analyzed, four email lists were selected for thorough observation, making it easier to infer general properties. These lists, picked as representing both a diverse set and usual lists, are:
\begin{itemize}
    \item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.}, with participants holding hybrid artistic and technological interests, from different countries. English is the language used the most. Abbreviated as LAU from now on.
    \item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.}, with participants are from different countries, and English is the language used the most. A more technical and less active version of LAU. Abbreviated LAD from now on.
    \item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.}, with computer programmers from different countries. English is the language used the most. Abbreviated as CPP from now on.
    \item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.}, with Brazilian activists holding digital culture interests. Portuguese is the most used language, although Spanish and English are also incident. Abbreviated MET from now on.
\end{itemize}

@@

\begin{table}
  \centering
  \caption{Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
$N$ is the number of participants (number of different email addresses).
$\Gamma$ is the number of threads (count of messages without antecedent).
$\overline{M}$ is the number of messages missing in the 20,000 collection, $100\frac{23}{20000}=0.115$ percent in the worst case.
A relation holds for all lists carefully considered: as the number of participants increases, the number of threads decreases.
This underpins a typology sketch of networks, as exposed in Section~\ref{sec:pty}.}
  \label{tab:genLists}
  \begin{tabular}{|l|c|c|c|c|c|}\hline
list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
	  \input{tables/tab1Geral}
  \end{tabular}
\end{table}

\section{Characterization methods}\label{sec:carac}
The email lists and the networks generated from them were characterized by: 1) statistics of activity along time, from seconds to years; 2) sectioning of the networks in hubs, intermediary and periphery vertices; 3) topological measures prominence regarding dispersion; 4) iterative visualization and data mining; 5) typological speculation about networks and participants.
Each of these procedures are described below.

Distribution of activity among participants received little dedicated attention, but is considered through almost all results of this article, as degree and strength are highly correlated to activity.

\subsection{Time activity statistics}\label{sec:mtime}
  Messages were counted along time with respect to seconds, minutes, hours, days of the week, days of the month, and months of the year. This resulted in histograms from which obvious patterns can be observed. Furthermore, the ration $\frac{b_h}{b_l}$ of the highest and lowest incidences on the histograms served as a hint of how the distribution relates to the uniform distribution.

  Measures of mean and dispersion were taken using circular statistics. For such, each $measure$ (data point) is represented as a complex number of unit magnitude $z=e^{i\theta}=\cos(\theta)+i\sin(\theta)$, where $\theta=measure\frac{2\pi}{period}$. Moments $m_n$, lengths of moments $R_n$, mean angle $\theta_\mu$, and rescaled mean angle $\theta_\mu'$ are defined as:

@@

The $\theta_\mu'$ is used as the measure of location. Dispersion is measured using the circular variance $Var(z)$, the circular standard deviation $S(z)$, and the circular dispersion $\delta(z)$:

@@

Although very simple, such measures enables both comparison of the email networks exposed in this article, and the deliverance of a benchmark to which other interaction networks can be compared.

\subsection{Interaction network}\label{intNet}
Interaction networks can be modeled both weighted or unweighted, both directed or undirected~\cite{bird,newmanCommunityDirected,newmanCommunity2013}.
Networks in this article are directed and weighted, the more informative of trivial possibilities, i.e. we did not observe directed unweighted, undirected weighted, and undirected unweighted representations of the interaction networks. 
The networks were obtained as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he/she read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network: B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This article uses the information network as described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to the edge weight. Self-loops were regarded as non-informative and discarded. These human social interaction networks are reported in the literature as exhibiting scale-free and small world properties, as expected for (some) social networks~\cite{bird,newmanBook}.

@@

Edges can be created from all antecedent message authors on the message-response thread to each message author.
We only linked the immediate antecedent to the new message author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a weaker connection between $x$ and $z$. Potential interpretations for this weaker connection are: double length, half weight or with one more ``obstacles''. This suggests pertinence of centrality measures that account for the connectivity with all nodes, such as betweenness centrality and accessibility~\cite{luMeasures,access}.

%\subsubsection{Sectioning of networks in periphery, intermediary and hubs sectors}\label{sectioning}
\subsection{Erd\"os sectioning}\label{sectioning}
In scale-free networks, the periphery, intermediary and hubs sectors can be derived from comparison with an Erd\"os-R\'enyi network with the same number of edges and vertices~\cite{3setores}, as depicted in Figure~\ref{fig:setores}. As a shorthand, this procedure is from now on called \emph{Erd\"os sectioning}, and the resulting sectors the \emph{Erd\"os sectors} or \emph{primitive sectors}.

@@

If $\mathcal{N}_f$ is directed and has no self-loops, the probability
of an edge between two arbitrary vertices is $p_e=\frac{z}{N(N-1)}$ (see Appendix~\ref{ap:ded}).
A vertex in the ideal Erd\"os-R\'enyi digraph with the same number of vertices and edges, and thus the same probability $p_e$ for the presence of an edge, will have degree $k$ with probability:

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/fser_}
    \caption{Degree distribution of scale-free and Erd\"os-R\'enyi ideal networks. The latter has more
        intermediary vertices, while the former has more peripheral and hub vertices. Sector borders are
        given by the two intersections $k_\Leftarrow$ and $k_\Rightarrow$ of the connectivity distributions. Characteristic degrees
    are in compact intervals of degree: $[0,k_\Leftarrow]$, $(k_\Leftarrow,k_\Rightarrow]$, $(k_\Rightarrow,k_{max}]$ for the Erd\"os sectors (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

@@

\noindent where \emph{way} can be \emph{in} or \emph{out}. In and out strengths ($s^{in}$, $s^{out}$) are divided by $\overline{w}$ and compared also using $\hat{P}$. Notice that $p_e$ remains the same, as each edge yields an incoming (or outgoing) edge, and there are at most $N(N-1)$ incoming (or outgoing) edges, thus $p_e=\frac{z}{N(N-1)}$ as with total degree.

@@

\begin{itemize}
    \item Exclusivist criterion $C_1$:  vertices are only classified if the class is the same according to all metrics. In this case, vertices classified (usually) does not reach 100\%, which is indicated by a black line in the figures of Appendix~\ref{figures}.
    \item Inclusivist criterion $C_2$: a vertex has the class given by any of the metrics. Therefore, a vertex may belong to more than one class, and total members may add more than 100\%, which is indicated by a black line in the figures of Appendix~\ref{figures}.
    \item Exclusivist cascade $C_3$: vertices are only classified as hubs if they are hubs according to all metrics. Intermediary are the vertices classified either as intermediary or hubs with respect to all metrics. The remaining vertices are regarded as peripheral.
    \item Inclusivist cascade $C_4$: vertices are hubs if they are classified as so according to any of the metrics. The remaining vertices are classified as intermediary if they belong to this category for any of the metrics. Peripheral vertices will then be those which were not classified as hub or intermediary with any of the metrics. 
    \item Exclusivist externals $C_5$: vertices are only hubs if they are classified as such according to all the metrics. The remaining vertices are classified as peripheral if they fall into the periphery or hub classes by any metric. The rest of the nodes are classified as intermediary.
    \item Inclusivist externals $C_6$: hubs are vertices classified as hubs according to any metric. The remaining vertices will be peripheral if they are classified as such according to any metric. The rest of the vertices will be intermediary vertices.
\end{itemize}

@@

The simplification of all the compound possibilities to the small set listed above can be formalized in strict mathematical terms, but this was considered out of the scope for current interests. Notice that the exclusivist cascade is the same sectioning that an inclusivist cascade from periphery to hubs (with inverted order of sectors precedence). Compound criteria can be used to examine network sections in the case of a low number of messages, such as in the last figures of Appendix~\ref{figures}.

Results from applying the Erd\"os sectioning, i.e. the achievement of well defined sectors by comparison of the real network with the Erd\"os R\'enyi model, are reported in Section~\ref{subsec:pih} and Appendix~\ref{figures}.

\subsection{Topology metrics for Principal Component Analysis}\label{measures}

The topology of the networks was observed by Principal Component Analysis (PCA~\cite{pca}) outcomes
with a small selection of the most basic and fundamental measurements for each vertex:

\begin{itemize}
    \item Degree     $k_i$: number of edges linked to vertex $i$.
    \item In-degree  $k_i^{in}$: number of edges ending at vertex $i$.
    \item Out-degree $k_i^{out}$: number of edges departing from vertex $i$.
    \item Strength $s$: sum of weights of all edges linked to vertex $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at vertex $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from vertex $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked.  The standard clustering coefficient for undirected graphs was used.
    \item Betweenness centrality $bt_i$: fraction of geodesics that contain the vertex $i$. Betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

In order to capture symmetries in the activity of participants, the following metrics were introduced for a vertex $i$ (see Section~\ref{prevalence}):

@@

\subsection{Evolution of the networks}
   The evolution of the networks was observed within a fixed number of messages, the window size $ws$. This same number of contiguous messages is considered with different shifts in the message timeline to obtain snapshots. Each snapshot was used to both do the Erd\"os sectioning (see Section~\ref{sectioning}) and to perform PCA with topological measures (see Section~\ref{measures}). Furthermore, this approach to evolution was used for visualization (see Section~\ref{sec:viz}) and for typological speculations (see Section~\ref{subsec:typ}).
The $ws$ used were 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000. Within a same $ws$, the number of vertices and edges vary in time, as do other network characteristics.

\subsection{Visualization of network evolution}\label{sec:viz}
The evolution of the networks was visualized with animations, image galleries and online gadgets made for this research~\cite{animacoes,galGMANE,appGMANE}. Such visualizations were crucial to guide research into the most important features of network evolution, and prompted us to capture the prominence of topological metrics along time using mean and standard deviations (see Section~\ref{measures} and Appendix~\ref{sec:pcat}), in addition to the size of the three sectors in a timeline fashion (Appendix~\ref{figures}). Visualization of network structure was specially useful as part of the email lists data mining, from which parts of relevant structures and results were driven. See Appendix~\ref{sec:related} for further directions about visualization and text mining concerning the results herein presented, with dedicated articles.

\subsection{Typology speculation}\label{subsec:typ}
Qualitative, typological speculations were the result of all the methodology. More specifically, the Erd\"os sectors was regarded as yielding a \emph{primitive typology} of human agents in social contexts with at least dozens of participants. 
Visualization and data mining were efficient in fueling speculations about the qualities of each type. 
The stability found in principal components of topological measures suggested that this primitive typology is general for (practically) all human contexts.
Also, the first author is a developer with cultural interests and has been directly and indirectly part of these communities for more than a decade.  This gave further safety about assumptions although care was taken not to regard this acquaintanceship as primary source or as definitive evidence.
The interested reader should see Appendix~\ref{ap:typ} for further context, 
as it might be considered audacious to bridge from a physics-based, 
quantitative classification to a qualitative and speculative approach.

@@

Remarkable features from the analysis of the four email lists are:
\begin{itemize}
    \item The activity along time is practically the same for all lists (Section~\ref{constDisc}) and suggests stable patterns.
    \item The fraction of participants in each Erd\"os sector is stable along time and can be observed even with very few messages (Section~\ref{subsec:pih}).
    \item The topological measures (regarded as most basic and still representative for a first analysis) combine into principal components in the same way for all lists and all snapshots. 
	    Furthermore, symmetry related measures, defined in this article, present more dispersion than the usual clustering coefficient (Section~\ref{prevalence}).
    \item Typology speculations are immediate from results (Section~\ref{sec:pty}).
\end{itemize}

The reminder of this section contains a detailed discussion of these results and Support Information holds further tables and figures from which such results can be better observed.

\subsection{Constancy and discrepancy of activity along time}\label{constDisc}

\begin{table*}[t]
	\caption{The rescaled circular mean $\theta_\mu'$, the standard deviation $S(z)$, the variance $Var(z)$, the circular dispersion $\delta(z)$ and the relation of maximum and minimum incidence at each time unit $\frac{max(incidence}{min(incidence}$. Also, $ \mu_{\frac{max(incidence')}{min(incidence')}} $ and $ \sigma_{\frac{max(incidence')}{min(incidence')} }$ are given for 1000 uniform distribution simulations within the same number of bins and with the same number of samples.
Section~\ref{sec:mtime} exposes the theoretical background considered of directional (or circular) statistics.}
\begin{center}
    \begin{tabular}{ |l|| c|c|c|c|c||c|c| }
        \hline
scale & $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$ & $\frac{max(incidence)}{min(incidence)}$ & $ \mu_{\frac{max(incidence')}{min(incidence')}} $ & $ \sigma_{\frac{max(incidence')}{min(incidence')} } $ \\ \hline\hline
	\input{tables/tab2TimeLAD}
    \end{tabular}
\end{center}
\label{tab:circ}
\end{table*}

One remarkable feature from the analysis is that the activity 
along time is practically the same for all lists. Circular mean and standard deviation
for each time scale are given in Table~\ref{tab:circ}.

Time scale also revealed patterns through histograms as exposed in Tables~\ref{tab:hin},~\ref{tab:win},~\ref{tab:min} and~\ref{tab:min2}. Seconds, minutes and years scales received the dedicated Sections~\ref{sec:secmin} and~\ref{sec:years} instead of tables.

\subsubsection{Seconds and minutes}\label{sec:secmin}
Messages were slightly more evenly distributed in all lists than in simulations using uniform distribution\footnote{Numpy version 1.6.1, ``random.randint'' function, was used for simulations, algorithms in \url{https://pypi.python.org/pypi/gmane}.}: for both seconds and minutes $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$ in the lists. Simulations reach these values but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.
Therefore, the incidence of messages at each second of a minute and at each minute of an hour was considered uniform, i.e. no trend was detected and the pattern is the uniformity. Circular dispersion is maximized and mean has little meaning.

>>>>

\section{Introduction}\label{sec:into}
Studies on human interaction networks have started long before modern computers, dating back to the nineteenth century, while the foundation of
social network analysis is generally attributed to the psychiatrist Jacob Moreno~\cite{newmanBook}. With the increasing availability of data related to human interactions, research on these networks has grown continuously. Contributions can now be found in a variety of fields in the literature, from social sciences and humanities~\cite{latour2013} to computer science~\cite{bird} and physics~\cite{barabasiHumanDyn,newmanFriendship}, given the multidisciplinary nature of the topic. One of the approaches from an exact science perspective is to represent interaction networks as complex networks [gmane,barabasiHumanDyn,newmanFriendship], with which 
several features of human interaction have been revealed. For example, the topology of human interaction networks exhibits a scale-free trace, which points to the existence of a small number of highly connected hubs and a large number of poorly connected nodes. The dynamics of complex networks representing human interaction has also been addressed ~\cite{barabasiEvo,newmanEvolving}, but only to a limited extent, since research is normally focused on a particular metric or functionality, such as accessibility or community detection~\cite{access,newmanModularity}.

In this paper we analyze the evolution of human interaction networks, by considering interaction in email lists as their representative. Using a timeline of activity snapshots with a constant number of contiguous messages in email lists, we found a remarkable stability for several of the network properties. Because this stability was shared by all email lists, we advocate that some of the conclusions can be valid for more general classes of interaction networks. In particular, this allows us to discuss typologies in the context of such networks, in an attempt to bridge the gap between approaches based solely on data analysis (i.e. from a hard sciences perspective) and those relevant to the social sciences. This is important insofar as typologies are the canon of scientific literature for classification of human agents~\cite{typCanon}.

The paper is organized as follows. Section~\ref{sec:related} describes related work, while details of the data and methods of analysis are given in  Section~\ref{sec:data} and Section~\ref{sec:carac}. Section~\ref{sec:results} brings the results and discussion, leading to Section~\ref{sec:conc} for conclusions and further work.

@@

The evolution of human interaction networks was addressed with a community focus, where the direction of edges was not taken into account~\cite{barabasiEvo}. Two topologically different networks emerged from human interaction networks, depending on the frequency of interactions, which can either be a generalized power law or an exponential connectivity distribution~\cite{barabasiTopologicalEv}. In email list networks, scale-free properties were reported with $\alpha=1$~\cite{bird} (as are web browsing and library loans~\cite{barabasiHumanDyn}), and different linguistic traces were related to weak and strong ties~\cite{GMANE2}.

Unreciprocated edges often exceed 50\% in the networks analyzed, which matches empirical evidence from the literature~\cite{newmanEvolving}. No correlation of topological characteristics and geographical coordinates was found~\cite{barabasiGeo}, therefore geographical positions were not considered in our study. Gender related behavior in mobile phone datasets has been reported~\cite{barabasiSex}, but this was not considered in the present article because email messages and addresses have no gender related metadata~\cite{GMANE}.

@@

We analyzed many email lists, but selected only four in order to make a thorough analysis, from which general properties can be inferred. These lists, selected as representing both a diverse set and ordinary lists, are:
\begin{itemize}
    \item Linux Audio Users list\footnote{gmane.linux.audio.users is list ID in GMANE.}, with participants holding hybrid artistic and technological interests, from different countries. English is the language used the most. Abbreviated as LAU from now on.
    \item Linux Audio Developers list\footnote{gmane.linux.audio.devel is list ID in GMANE.}, with participants from different countries, and English is the language used the most. A more technical and less active version of LAU. Abbreviated LAD from now on.
    \item Development list for the standard C++ library\footnote{gmane.comp.gcc.libstdc++.devel is list ID in GMANE.}, with computer programmers from different countries. English is the language used the most. Abbreviated as CPP from now on.
    \item List of the MetaReciclagem project\footnote{gmane.politics.organizations.metareciclagem is list ID in GMANE.}, with Brazilian activists holding digital culture interests. Portuguese is the most used language, although Spanish and English are also incident. Abbreviated MET from now on.
\end{itemize}

@@

\begin{table}
  \centering
  \caption{Columns $date_1$ and $date_M$ have dates of first and last messages from the 20,000 messages considered in each email list.
$N$ is the number of participants (number of different email addresses).
$\Gamma$ is the number of threads (count of messages without antecedent).
$\overline{M}$ is the number of messages missing in the 20,000 collection, $100\frac{23}{20000}=0.115$ percent in the worst case.
A relation holds for all lists carefully considered: as the number of participants increases, the number of threads decreases.
This underpins a typology sketch of networks, as discussed in Section~\ref{sec:pty}.}
  \label{tab:genLists}
  \begin{tabular}{|l|c|c|c|c|c|}\hline
list & $date_1$ & $date_{M}$    & $N$  & $\Gamma$ & $\overline{M}$ \\\hline
	  \input{tables/tab1Geral}
  \end{tabular}
\end{table}

\section{Characterization methods}\label{sec:carac}
The email lists and the networks generated from them were characterized by using five procedures, namely: 1) statistics of activity along time, from seconds to years; 2) sectioning of the networks in hubs, intermediary and peripheral vertices; 3) topological metrics and their dispersion; 4) iterative visualization and data mining; 5) typological speculation about networks and participants.
Each of these procedures are described below.

\subsection{Time activity statistics}\label{sec:mtime}
  Messages were counted along time with respect to seconds, minutes, hours, days of the week, days of the month, and months of the year. This resulted in histograms from which patterns could be drawn. The ratio $\frac{b_h}{b_l}$ between the highest and lowest incidences on the histograms served as a hint of how the observed distribution is compared to a uniform distribution.

  The average and the dispersion were taken using circular statistics, in which each $measurement$ (data point) is represented as a complex number with modulus equal to one, $z=e^{i\theta}=\cos(\theta)+i\sin(\theta)$, where $\theta=measurement\frac{2\pi}{period}$. The moments $m_n$, lengths of moments $R_n$, mean angle $\theta_\mu$, and rescaled mean angle $\theta_\mu'$ are defined as:

@@

$\theta_\mu'$ is used as the measure of location. Dispersion is measured using the circular variance $Var(z)$, the circular standard deviation $S(z)$, and the circular dispersion $\delta(z)$:

@@

\subsection{Interaction networks}\label{intNet}
Interaction networks can be modeled both weighted or unweighted, both directed or undirected~\cite{bird,newmanCommunityDirected,newmanCommunity2013}.
Networks in this article are directed and weighted, the more informative of trivial possibilities, i.e. we did not investigate directed unweighted, undirected weighted, and undirected unweighted representations of the interaction networks. 
The networks were obtained as follows: a direct response from participant B to a message from participant A yields an edge from A to B, as information went from A to B. The reasoning is: if B wrote a response to a message from A, he/she read what A wrote and formulated a response, so B assimilated information from A, thus $A \rightarrow B$. Inverting edge direction yields the status network: B read the message and considered what A wrote worth responding, giving status to A, thus $B\rightarrow A$. This article uses the information network as described above and depicted in Figure~\ref{formationNetwork}. Edges in both directions are allowed. Each time an interaction occurs, one is added to the edge weight. Self-loops were regarded as non-informative and discarded. These human social interaction networks are reported in the literature as exhibiting scale-free and small world properties, as expected for (some) social networks~\cite{bird,newmanBook}.

@@

Edges can be created from all antecedent message authors on the message-response thread to each message author.
We only linked the immediate antecedent to the new message author, both for simplicity and for the valid objection that in adding two edges, $x\rightarrow y$ and $y\rightarrow z$, there is also a weaker connection between $x$ and $z$. Potential interpretations for this weaker connection are: double length, half weight or with one more ``obstacles''. This suggests the adequacy of centrality measurements to account for the connectivity with all nodes, such as betweenness centrality and accessibility~\cite{luMeasures,access}.

%\subsubsection{Sectioning of networks in peripheral, intermediary and hubs sectors}\label{sectioning}
\subsection{Erd\"os sectioning}\label{sectioning}
In scale-free networks, the peripheral, intermediary and hubs sectors can be derived from a comparison with an Erd\"os-R\'enyi network with the same number of edges and vertices~\cite{3setores}, as depicted in Figure~\ref{fig:setores}. We shall refer to this procedure as \emph{Erd\"os sectioning}, with the resulting sectors being referred to as \emph{Erd\"os sectors} or \emph{primitive sectors}.

@@

If $\mathcal{N}_f$ is directed and has no self-loops, the probability
of an edge between two arbitrary vertices is $p_e=\frac{z}{N(N-1)}$ (see Appendix~\ref{ap:ded}). (a numeração de figuras, tabelas, apêndices, etc. tem que começar de 1 ou A)
A vertex in the ideal Erd\"os-R\'enyi digraph with the same number of vertices and edges, and thus the same probability $p_e$ for the presence of an edge, will have degree $k$ with probability:

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{figs/fser_}
    \caption{Degree distribution of scale-free and Erd\"os-R\'enyi ideal networks. The latter has more
        intermediary vertices, while the former has more peripheral and hub vertices. The sector borders are defined by the two intersections $k_\Leftarrow$ and $k_\Rightarrow$ of the connectivity distributions. Characteristic degrees
    are in compact intervals of degree: $[0,k_\Leftarrow]$, $(k_\Leftarrow,k_\Rightarrow]$, $(k_\Rightarrow,k_{max}]$ for the Erd\"os sectors (periphery, intermediary and hubs).}
    \label{fig:setores}
\end{figure}

@@

\noindent where \emph{way} can be \emph{in} or \emph{out}. In and out strengths ($s^{in}$, $s^{out}$) are divided by $\overline{w}$ and compared also using $\hat{P}$. Note that $p_e$ remains the same, as each edge yields an incoming (or outgoing) edge, and there are at most $N(N-1)$ incoming (or outgoing) edges, thus $p_e=\frac{z}{N(N-1)}$ as with the total degree.

@@

\begin{itemize}
    \item Exclusivist criterion $C_1$:  vertices are only classified if the class is the same according to all metrics. In this case, vertices classified (usually) do not reach 100\%, which is indicated by a black line in Figures~\ref{fig:sectIL}.
    \item Inclusivist criterion $C_2$: a vertex has the class given by any of the metrics. Therefore, a vertex may belong to more than one class, and total members may add more than 100\%, which is indicated by a black line in Figure~\ref{fig:sectIL}.
    \item Exclusivist cascade $C_3$: vertices are only classified as hubs if they are hubs according to all metrics. Intermediary are the vertices classified either as intermediary or hubs with respect to all metrics. The remaining vertices are regarded as peripheral.
    \item Inclusivist cascade $C_4$: vertices are hubs if they are classified as so according to any of the metrics. The remaining vertices are classified as intermediary if they belong to this category for any of the metrics. Peripheral vertices will then be those which were not classified as hub or intermediary with any of the metrics. 
    \item Exclusivist externals $C_5$: vertices are only hubs if they are classified as such according to all the metrics. The remaining vertices are classified as peripheral if they fall into the periphery or hub classes by any metric. The rest of the nodes are classified as intermediary.
    \item Inclusivist externals $C_6$: hubs are vertices classified as hubs according to any metric. The remaining vertices will be peripheral if they are classified as such according to any metric. The rest of the vertices will be intermediary vertices.
\end{itemize}

@@

The simplification of all the compound possibilities to the small set listed above can be formalized in strict mathematical terms, but this was considered out of the scope for current interests. It is worth noting that the exclusivist cascade is the same sectioning of an inclusivist cascade from periphery to hubs, but with inverted order of sectors precedence. These compound criteria can be used to examine network sections in the case of a low number of messages, such as in the last figures of Support Information.

\subsection{Topological metrics for Principal Component Analysis}\label{measures}

The topology of the networks was studied using Principal Component Analysis (PCA~\cite{pca}) with a small selection of the most basic and fundamental measurements for each vertex, as follows:

\begin{itemize}
    \item Degree     $k_i$: number of edges linked to vertex $i$.
    \item In-degree  $k_i^{in}$: number of edges ending at vertex $i$.
    \item Out-degree $k_i^{out}$: number of edges departing from vertex $i$.
    \item Strength $s$: sum of weights of all edges linked to vertex $i$.
    \item In-strength $s_i^{in}$: sum of weights of all edges ending at vertex $i$.
    \item Out-strength $s_i^{out}$: sum of weights of all edges departing from vertex $i$.
    \item Clustering coefficient $cc_i$: fraction of pairs of neighbors of $i$ that are linked.  The standard clustering coefficient for undirected graphs was used.
    \item Betweenness centrality $bt_i$: fraction of geodesics that contain vertex $i$. The betweenness centrality index considered directions and weight, as specified in~\cite{faster}.
\end{itemize}

In order to capture symmetries in the activity of participants, the following metrics were introduced for a vertex $i$:

@@

\subsection{Evolution of the networks}
   The evolution of the networks was observed within a fixed number of messages, which we refer to as the window size $ws$. This same number of contiguous messages is considered with different shifts in the message timeline to obtain snapshots. Each snapshot was used both to perform the Erd\"os sectioning and apply PCA for the topological metrics.  
The $ws$ used were 50, 100, 200, 400, 500, 800, 1000, 2000, 2500, 5000 and 10000. Within a same $ws$, the number of vertices and edges vary in time, as do other network characteristics. Such changes could be visualized with the tools described below.

\subsection{Visualization of network evolution}\label{sec:viz}
The evolution of the networks was visualized with animations, image galleries and online gadgets developed specifically for this research~\cite{animacoes,galGMANE,appGMANE}. Such visualizations were crucial to guide research into the most important features of network evolution, and prompted us to capture the prominence of topological metrics along time using mean and standard deviations. Furthermore, the size of three sectors could be visualized in a timeline fashion (Figure~\ref{fig:sectIL}). Visualization of network structure was especially useful as part of the email lists data mining, from which parts of relevant structures and results were driven.

\subsection{Typology speculation}\label{subsec:typ}
(queria discutir pessoalmente este ponto)
Qualitative, typological speculations were the result of all the methodology. More specifically, the Erd\"os sectors was regarded as yielding a \emph{primitive typology} of human agents in social contexts with at least dozens of participants. 
Visualization and data mining were efficient in fueling speculations about the qualities of each type. 
The stability found in principal components of topological measures suggested that this primitive typology is general for (practically) all human contexts.
Also, the first author is a developer with cultural interests and has been directly and indirectly part of these communities for more than a decade.  This gave further safety about assumptions although care was taken not to regard this acquaintanceship as primary source or as definitive evidence.
The interested reader should see Appendix~\ref{ap:typ} for further context, 
as it might be considered audacious to bridge from a physics-based, 
quantitative classification to a qualitative and speculative approach.

@@

Remarkable features from the analysis of the four email lists are:
\begin{itemize}
    \item The activity along time is practically the same for all lists, thus suggesting stable patterns.
    \item The fraction of participants in each Erd\"os sector is stable along time and can be determined even with very few messages 
    \item The topological metrics combine into principal components in PCA in the same way for all lists and all snapshots (). 
        \item Symmetry measures of the topology, as defined in this article, present more dispersion than the usual clustering coefficient (Section~\ref{prevalence}).
    \item Typology speculations are immediate from results (Section~\ref{sec:pty}).
\end{itemize}

\subsection{Activity along time}\label{constDisc}

\begin{table*}[t]
	\caption{The rescaled circular mean $\theta_\mu'$, the standard deviation $S(z)$, the variance $Var(z)$, the circular dispersion $\delta(z)$ and the relation of maximum and minimum incidence at each time unit $\frac{max(incidence}{min(incidence}$. Also, $ \mu_{\frac{max(incidence')}{min(incidence')}} $ and $ \sigma_{\frac{max(incidence')}{min(incidence')} }$ are given for 1000 uniform distribution simulations within the same number of bins and with the same number of samples. Section~\ref{sec:mtime} describes the theoretical background of directional (or circular) statistics. This typical table was made using LAD list messages.}
\begin{center}
    \begin{tabular}{ |l|| c|c|c|c|c||c|c| }
        \hline
scale & $\theta_\mu'$ & $S(z)$ & $Var(z)$ & $\delta(z)$ & $\frac{max(incidence)}{min(incidence)}$ & $ \mu_{\frac{max(incidence')}{min(incidence')}} $ & $ \sigma_{\frac{max(incidence')}{min(incidence')} } $ \\ \hline\hline
	\input{tables/tab2TimeLAD}
    \end{tabular}
\end{center}
\label{tab:circ}
\end{table*}

The activity along time is practically the same for all lists, as exemplified with the circular statistics in Table~\ref{tab:circ} and Support information.

\subsubsection{Seconds and minutes}\label{sec:secmin}
Messages were slightly more evenly distributed in all lists than in simulations using uniform distribution\footnote{Numpy version 1.6.1, ``random.randint'' function, was used for simulations, algorithms in \url{https://pypi.python.org/pypi/gmane}.}: for both seconds and minutes $\frac{max(incidence)}{min(incidence)} \in (1.26,1.275]$ in the lists. Simulations reach these values but have in average more discrepant higher and lower peaks $\xi=\frac{max(incidence')}{min(incidence')} \Rightarrow \mu_\xi=1.2918 \text{ and } \sigma_\xi=0.04619$.
Therefore, the incidence of messages at each second of a minute and at each minute of an hour was considered uniform, i.e. no trend was detected and the pattern is the uniformity. Circular dispersion is maximized and the mean has little meaning.
